---
title: "Fraud Prediction using Decision Tree"
author: 2324244
output: html_notebook
---

# Loading necessary libraries for the project
```{r}
install.packages('dplyr')
install.packages('ggplot2')
install.packages('tree')
install.packages('randomForest')
install.packages('rpart')
install.packages('rpart.plot')
install.packages('pROC')
library(dplyr)
library(ggplot2)
library(readr)
library(rpart)
library(tree)
library(randomForest)
library(rpart)
library(rpart.plot)
library(pROC)
```

# Reading the fraud dataset from a CSV file
```{r}
fraud_data <- read.csv('Fraud_Data1.csv', stringsAsFactors = T)
```

# Examine the structure of the dataset and display the first five rows
```{r}
str(fraud_data)

head(fraud_data, 5)
```

# Visualize the distribution of transaction amounts using a histogram
```{r}
ggplot(fraud_data, aes(x = amount)) + 
  geom_histogram(bins = 50) + scale_x_log10() + 
  ggtitle("Distribution of Transaction Amounts")
```

# Creating a bar chart to show the average transaction amount by transaction type
```{r}
bar_chart <- ggplot(fraud_data,aes(x=type,y=amount,fill=type))+
  geom_bar(stat = "summary", fun = "mean")+
  labs(title = "Average amount by Transaction type", x="Trasaction Type", y="Amount")+
  theme_minimal()
print(bar_chart)
```

# Calculate and visualize average transaction amounts for fraud and non-fraud transactions
```{r}
average_fraud <- mean(fraud_data$amount[fraud_data$isFraud == 1])
average_notfraud <- mean(fraud_data$amount[fraud_data$isFraud == 0])

average_data <- data.frame(Category = c("Fraud", "Not Fraud"),
                           Average = c(average_fraud, average_notfraud))

print(average_data)

ggplot(average_data, aes(x = Category, y = Average, fill = Category)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Fraud" = "red", "Not Fraud" = "blue")) + 
  labs(title = "Average Fraud vs Not Fraud", x = "Category", y = "Average") +
  theme_minimal()
```

# Show the total number of fraudulent transactions by type
```{r}
sum_isFraud <- aggregate(isFraud ~ type, data = fraud_data, FUN = sum)
ggplot(sum_isFraud,aes(x = type, y = isFraud))+
  geom_bar(stat = "identity", fill = "red")+
  labs(title = "Number of Fraud Transaction by Type", x="Trasaction Type", y="Sum of isFraud")+
  theme_minimal()
```

# Histogram to show transaction amount distribution by fraud status
```{r}
ggplot(fraud_data, aes(x = amount, fill = as.factor(isFraud))) + 
  geom_histogram(bins = 50, position = "identity", alpha = 0.5) + 
  scale_x_log10() + 
  ggtitle("Transaction Amount by Fraud Status")
```

# Pie chart showing the distribution of transaction types as a percentage
```{r}
per_trans_type <- prop.table(table(fraud_data$type))*100

plot_fuel <- data.frame(
  type = names(per_trans_type),
  percentage = as.numeric(per_trans_type))

ggplot(plot_fuel,aes(x="",y = percentage, fill=type,group=type))+
  geom_bar(width = 1, stat = "identity",color = "black")+
  geom_text(aes(label = paste0(round(percentage),"%")),position = position_stack(vjust = 0.5))+
  coord_polar(theta = "y")+
  labs(title = "Distribution of Transaction Type",fill="Transaction Type")+
  theme_void()
```

# Pie chart showing the count of fraudulent versus non-fraudulent transactions
```{r}
fraud_count <- table(fraud_data$isFraud)

count_fraud_data <- data.frame(Category = names(fraud_count), Count = as.numeric(fraud_count))

ggplot(count_fraud_data , aes(x = "", y = Count, fill = Category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y")+
  labs(title = "Distribution of Transaction Type")+
  scale_fill_manual(values = c("0" = "skyblue", "1" = "Red"))+
  theme_void()+
  theme(legend.position = "bottom")
```

# Data Preparation and Balancing

```{r}
# Separate the fraud and non-fraud transactions into two different data frames
fraud_transactions <- fraud_data %>% filter(isFraud == 1)
non_fraud_transactions <- fraud_data %>% filter(isFraud == 0)

# Perform undersampling on the non-fraud transactions to match the number of fraud transactions
set.seed(123) 
non_fraud_undersampled <- non_fraud_transactions %>% sample_n(nrow(fraud_transactions))

# Combine the undersampled non-fraud transactions with the fraud transactions to create a balanced dataset
balanced_df <- rbind(fraud_transactions, non_fraud_undersampled)

# balanced_df now contains an equal number of fraud and non-fraud transactions
```

# Split Data into Training and Test Sets
```{r}
set.seed(123)

# Assuming balanced_df is your balanced dataset
indices <- sample(1:nrow(balanced_df), size = 0.7 * nrow(balanced_df))
train_data <- balanced_df[indices, ]
test_data <- balanced_df[-indices, ]
```

# Model Building
```{r}
# Fit the model
model <- rpart(isFraud ~ step + type + amount + oldbalanceOrg + newbalanceOrig, data = train_data, method = "class")

# Plot the decision tree
rpart.plot(model, main="Decision Tree for Fraud Detection")
```

# Model Evaluation
```{r}
# Make predictions
predictions <- predict(model, test_data, type = "class")

# Evaluate model performance
confusionMatrix <- table(Predicted = predictions, Actual = test_data$isFraud)
print(confusionMatrix)

# You can also calculate accuracy, precision, recall, etc., based on the confusion matrix
```

# ROC Curve Analysis

```{r}
# Make predictions to get probabilities
# Assuming the second column of the prediction gives the probability for class "1"
predictions_prob <- predict(model, test_data, type = "prob")[,2]

# Calculate ROC curve
roc_result <- roc(test_data$isFraud, predictions_prob)

# Plot ROC curve
plot(roc_result, main="ROC Curve", col="#1c61b6")
# Add AUC to the plot
auc(roc_result)
print(paste("AUC:", auc(roc_result)))
```

